Anomaly detection with kernel density estimation on manifolds

### 2022-05-15 paper/paper.tex
True volume density function to be used, also try small N, optimize h with AMSE
Finalise example
Submit abstract

\begin{document}
\maketitle
\begin{abstract}
Kernel density estimation based on random samples often involves a crucial bandwidth selection. For data points lying on the Riemannian manifolds embedded in a high-dimensional space, manifold learning aims to find the low-dimensional representation of the data but there is local distortion for each observation, which failed to be captured by the kernel density estimation. We propose a new kernel density estimation of manifold learning embedding with the Riemannian metric being the pointwise bandwidth matrix and further identify outliers with the lowest densities. The Riemannian metric measures the direction and angle of the local distortion induced by the non-linear functions in manifold learning and is used for smoothing the density estimates. To compare our proposed method with fixed bandwidth KDE, we run two simulations with a 2-D meta data mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere meta data mapped in 100-D space, and show that the proposed estimator could improve the density estimates given a good manifold learning embedding and gives higher correlations between ranks of true density and estimated density. For application, we use the Irish smart meter data and explore the distributions of different households and periods of the week. Both density estimators are compared by examing the distributions of the most typical households with highest densities and the most anomalous households with lowest densities.
\end{abstract}
\begin{keywords}
manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels
\end{keywords}

@@

Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\). Let
\(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\)
is the orthonormal basis of \(\mathbb{R}^3\). Consider the exponential
chart \((U, \Psi)\) induced by the parametrisation
\(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by \[
\Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
\] if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where
\(r=\sqrt{s^2 + t^2}\). Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
Then we have \[
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)}, \\
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)}.
\] Then each component of the riemannian metric could be calculated as
\(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\). We obtain
that the volume density on the sphere is \[
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
\] for \(q \neq p, -p\), and \(\theta_p(p) = 1\).

In \(S^1\), polar coordinates: \(\phi(\exp_p{vt})^{\prime}(0) = v\)

cartesian coordinates:
\(\psi(\exp_p{vt})^{\prime}(0) = x_p \cos{v} - \sqrt{1-x_p^2} \sin{v}\)
\bigg\_t=0 = -\sqrt{1-x_p^2}v\$

>>>>

\begin{document}
\maketitle
\begin{abstract}
Manifold learning can be used to obtain a low-dimensional representation of the underlying manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the underlying Riemannian manifold. We propose a novel kernel density estimator for any manifold learning embedding by introducing the estimated Riemannian metric of the manifold as the variable bandwidth matrix for each point. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D metadata mapped into a 3-D swiss roll or twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space, and demonstrate that the proposed estimator could improve the density estimates given a good manifold learning embedding and has higher rank correlations between the true and estimated manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates our estimator's capability to fix the distortion of the manifold geometry and to be further used for anomaly detection in high-dimensional data.
\end{abstract}
\begin{keywords}
manifold learning, variable bandwidth, Riemannian metric, highest density region, Gaussian kernels
\end{keywords}

@@

Let \(S^2\) be the two-dimensional sphere of radius 1 and \(p \in S^2\). Let
\(v\) and \(w\) be vectors such that \(\|v\| = \|w\| = 1\) and \(\{v, w, p\}\)
is the orthonormal basis of \(\mathbb{R}^3\). Consider the exponential
chart \((U, \Psi)\) induced by the parametrisation
\(\Psi^{-1}: B_\pi(0) \rightarrow S^2 \backslash \{-p\}\) given by \[
\Psi^{-1}(s,t) = \cos(r)p + \sin(r)(sv+tw)r^{-1},
\] if \((s,t) \neq (0,0)\) and \(\Psi^{-1}(0,0)=p\) where
\(r=\sqrt{s^2 + t^2}\). Note that \(r=d_g(p,q)\) if \(q = \exp_p(sv+tw)\).
Then we take the partial derivatives of \$\textbackslash Psi\^{}\{-1\}(s,t)\$ with
respect to \$s\$ and \$t\$ and have that
\[
A(s,t) = \frac{\partial}{\partial \Psi_1} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}s}{r}p + \frac{\cos(r)rs^2+t^2\sin(r)}{r^3}v + \frac{(\cos(r)r-\sin(r))st}{r^3}w,
\]
\[
B(s,t) = \frac{\partial}{\partial \Psi_2} \bigg|_{\Psi^{-1}(s,t)} = \frac{-\sin{(r)}t}{r}p + \frac{(\cos(r)r-\sin(r))st}{r^3}v + \frac{\cos(r)rt^2+\sin(r)s^2}{r^3}w.
\]
Then each component of the riemannian metric could be calculated as
\(g(A(s,t), A(s,t)), g(A(s,t), B(s,t)), g(B(s,t), B(s,t))\). We obtain
that the volume density on the sphere is \[
\theta_p(q) = \bigg| \det g_q\bigg( \frac{\partial}{\partial \Psi_i}\bigg|_q, \frac{\partial}{\partial \Psi_j}\bigg|_q \bigg) \bigg|^{1/2} \\
= \frac{|\sin(d_g(p,q))|}{d_g(p,g)},
\] for \(q \neq p, -p\), and \(\theta_p(p) = 1\).

Similarly in \(S^1\), we could let \(\{v, p\}\) be the basis of
\(\mathbb{R}^2\). Then the exponential chart induced

Polar coordinates:
\(\phi(\exp_p{vt})^{\prime}(0) = (\alpha + vt)^\prime\bigg|_{t=0} = v\).

Cartesian coordinates:
\[
\psi(\exp_p{vt})^{\prime}(0) = (\cos(\arccos(p)+vt))^\prime\bigg|_{t=0} \\
= (p \cos{v} - \sqrt{1-p^2} \sin{v})^\prime \bigg|_{t=0} \\
= - \sqrt{1 - p^2}v
\]
Then the volume density function is
\[
\theta_p(q) = \sqrt{(-\sqrt{1-p^2})^2} = \sqrt{1-p^2},
\]
and the geodesic distance is
\[
d_g(p,q) = \arccos(p) - \arccos(q).
\]
